---
title: "Final Project Part 2"
author:
- "Maximilien Filion - 20598645, "
- "Aneesh Kaura - 66832304, "
- "Monica Mihailescu - 47409289 "
date: "2025-11-17"
output:
  word_document: default
  html_document: default
  pdf_document: default
---
# ============================================================================
# RNA-seq Analysis Template: Clustering and Differential Expression
# ============================================================================
# This template provides a structured framework for comprehensive gene 
# expression analysis from preprocessing through clinical validation
# ============================================================================

# Load required libraries
```{r, results='hide'}
#if (!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install("org.Hs.eg.db")
# BiocManager::install("AnnotationDbi") 

# Install clusterProfiler properly
#BiocManager::install("clusterProfiler", ask = FALSE, update = FALSE)

#BiocManager::install("DOSE")
#BiocManager::install("pathview") 

library(clusterProfiler)
library(AnnotationDbi)
library(DESeq2)
library(pheatmap)
library(ggplot2)
library(cluster)
library(survival)
library(survminer)
#library(clusterProfiler)
library(org.Hs.eg.db) # Contains human gene annotations
library(dplyr)
library(tibble)
library(biomaRt)   
```


set.seed(1)

# ============================================================================
# STEP 1: DATA LOADING AND INITIAL SETUP
# ============================================================================

# Load the data files
# - count_matrix: raw counts (genes x samples)
# - clinical_data: patient clinical information
# - mutation_data: mutation status for each patient
```{r}
#Load the data (from lab 7)
RNAseq_KIRC <- read.csv("RNAseq_KIRC.csv")
data_clinical_patient <- read.delim("data_clinical_patient.txt", header = TRUE, sep = "\t", skip = 4)
data_mutations <- read.delim("data_mutations.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE)
```

```{r}
head(RNAseq_KIRC)
head(data_clinical_patient)
```


# Creates a dataframe containing all the patients common to the three datasets.
will be used in downstream applications.
```{r}
# RNAseq_KIRC dataset
rna_patient_ids <- gsub("\\.", "-", substr(colnames(RNAseq_KIRC), 1, 12))
# Data_clinical_patient dataset
clinical_patient_ids <- unique(data_clinical_patient$PATIENT_ID)
# Data_mutation datasetd
mutation_patient_ids <- unique(substr(data_mutations$Tumor_Sample_Barcode, 1, 12))
# Common patients
common_patients <- Reduce(intersect, list(rna_patient_ids,
clinical_patient_ids, mutation_patient_ids))

# Patients in all three datasets
common_patients <- Reduce(intersect, list(
  rna_patient_ids,
  clinical_patient_ids,
  mutation_patient_ids
))
cat("Number of patients in ALL datasets:", length(common_patients), "\n")
```

# Trim the datasets to only include the common patients
```{r}
#For the RNA seq dataset, we want to keep all the common patient columns and the ensembl gene ID (i.e.Column "X")
# Trim the RNA sequencing dataset
common_patients_and_X <- c(common_patients, "X")
RNAseq_KIRC <- RNAseq_KIRC[ , gsub("\\.", "-", substr(colnames(RNAseq_KIRC), 1, 12)) %in% common_patients_and_X]

# Trim the clinical dataset
data_clinical_patient <- data_clinical_patient[data_clinical_patient$PATIENT_ID %in% common_patients, ]

# Trim the mutation dataset
data_mutations <- data_mutations[substr(data_mutations$Tumor_Sample_Barcode, 1, 12) %in% common_patients, ]
```


<!-- # Merge the different ensembl version in the RNAseq dataframe -->
<!-- ```{r} -->
<!-- # Strip the ensembl version -->
<!-- RNAseq_KIRC$X <- gsub("\\..*$", "", RNAseq_KIRC$X) -->

<!-- # Collapse the duplicates -- GenAI was used for this part -->
<!-- RNAseq_KIRC_novers <- RNAseq_KIRC %>% -->
<!--   group_by(X) %>% -->
<!--   summarise(across(everything(), sum)) -->
<!-- ``` -->
The code above only cuts ~50 duplicate genes and takes a while to run. It was commented to signify that we thought about it but did not use it due to computational ressources. 


# Check for missing values
```{r}
# RNA-seq:
rnaseq_missing <- sum(is.na(RNAseq_KIRC))
cat("Missing values in RNA-seq count matrix:", rnaseq_missing, "\n")

# Clinical: Check key columns
key_clinical_cols <- c("PATIENT_ID", "AGE", "SEX", "OS_STATUS", "OS_MONTHS", "DSS_STATUS", "DSS_MONTHS", "DFS_STATUS", "DFS_MONTHS", "PFS_STATUS", "PFS_MONTHS", "PATH_T_STAGE", "PATH_N_STAGE", "PATH_M_STAGE", "AJCC_PATHOLOGIC_TUMOR_STAGE")
existing_cols <- key_clinical_cols[key_clinical_cols %in% colnames(data_clinical_patient)]

if(length(existing_cols) > 0) {
  for(col in existing_cols) {
    missing_count <- sum(is.na(data_clinical_patient[[col]]) | data_clinical_patient[[col]] == "")
    cat("Missing/empty values in", col, ":", missing_count, "\n")
  }
}

# Mutations: Check key columns
cat("\nMissing values in Tumor_Sample_Barcode:", sum(is.na(data_mutations$Tumor_Sample_Barcode)), "\n")
cat("Missing/empty values in Hugo_Symbol:", sum(is.na(data_mutations$Hugo_Symbol) | data_mutations$Hugo_Symbol == ""), "\n")
cat("Missing/empty values in Variant_Classificatoin:", sum(is.na(data_mutations$Variant_Classification) | data_mutations$Variant_Classificationl == ""), "\n")
cat("Missing/empty values in Consequence:", sum(is.na(data_mutations$Consequence) | data_mutations$Consequence == ""), "\n")
cat("Missing/epmty values in Impact:", sum(is.na(data_mutations$IMPACT) | data_mutations$IMPACT == ""), "\n\n")
```


# ============================================================================
# STEP 2: PREPROCESSING
# ============================================================================

## 2.1: Normalization using DESeq2
# GOAL: Account for sequencing depth and composition bias

# TODO: Create DESeq2 object from raw counts
# TODO: Estimate size factors
# TODO: Extract normalized counts
```{r}
# Create DESeq2 object from raw counts
# Prepare count matrix (genes as rows, samples as columns)
count_matrix <- as.matrix(RNAseq_KIRC[, -1])  # Remove gene name column
rownames(count_matrix) <- RNAseq_KIRC[, 1]    # Set gene names as row names

# Create a simple sample metadata data frame
# For normalization, we just need a basic colData with sample names
coldata <- data.frame(
  sample = colnames(count_matrix),
  condition = rep("tumor", ncol(count_matrix))  # Placeholder condition
)
rownames(coldata) <- coldata$sample

cat("Count matrix dimensions:", dim(count_matrix), "\n")

# Create DESeq2 dataset object
dds <- DESeqDataSetFromMatrix(
  countData = count_matrix,
  colData = coldata,
  design = ~ 1  # No DE, normalization only
)
cat("DESeq2 object created successfully\n\n")

# Estimate size factors
dds <- estimateSizeFactors(dds)

# Get size factors
size_factors <- sizeFactors(dds)
cat("Size factor summary:\n")
print(summary(size_factors))
cat("\n")

# Extract normalized counts
normalized_counts <- counts(dds, normalized = TRUE)
cat("Normalized count matrix dimensions:", dim(normalized_counts), "\n")
```

## 2.2: Log-transformation
# GOAL: Stabilize variance and make data more normally distributed

# TODO: Apply log2 transformation and VST
```{r}
# simple Log2(x+1) transformation 
log_norm_counts <- log2(normalized_counts + 1)

cat("Log transform count matrix dimensions:", dim(log_norm_counts), "\n")
cat("Log transformation range:", range(log_norm_counts), "\n\n")

# Apply VST (recommended for >30 samples) - Applies variance-stabilizing transformation, 'blind' Ignores experimental design
vsd <- vst(dds, blind = TRUE)
vst_counts <- assay(vsd)

cat("VST count matrix dimensions:", dim(vst_counts), "\n")
cat("VST value range:", range(vst_counts), "\n\n")
```

## 2.3: Z-score transformation
# GOAL: Standardize gene expression across samples

# TODO: Calculate z-scores per gene (across samples)
# TODO: Verify mean ≈ 0 and sd ≈ 1 for each gene
```{r}
# Z-score the log-transformed counts (by gene)
z_log <- t(scale(t(log_norm_counts)))

cat("Z-scored log2 matrix dimensions:", dim(z_log), "\n")
cat("Z-scored log2 range:", range(z_log, na.rm = TRUE), "\n\n")

# Z-score the VST-transformed counts (by gene)
z_vst <- t(scale(t(vst_counts)))

cat("Z-scored VST matrix dimensions:", dim(z_vst), "\n")
cat("Z-scored VST range:", range(z_vst, na.rm = TRUE), "\n\n")
```


# ============================================================================
# INVESTIGATION 1: Global Expression-Based Clustering
# RQ 2.1: Do patients cluster into distinct phenotype subtypes?
# ============================================================================

# Filter the most variable genes 
Running distance on 60660 observations was too computationally intensive. We decided to only keep the top 3000 genes since PCA will explain only part of the variance and variance will be mostly explained by those highly variable 3000 genes. It was chosen arbitrarily 

```{r}
# compute variance on raw data
gene_vars <- apply(normalized_counts, 1, var)


#Rank the genes by variance
top_genes <- names(sort(gene_vars, decreasing = TRUE))[1:3000]

# Create a subset matrix with the top 3000 genes
expr_z_top <- z_vst[top_genes, ]
```


# Plot PCA and look for distinct group
```{r}
pca_res <- prcomp(t(expr_z_top), scale = TRUE)

# Percentage variance explained
pve <- (pca_res$sdev^2) / sum(pca_res$sdev^2)

# Scree plot
ggplot(data.frame(PC = 1:10, Var = pve[1:10]),
       aes(x = PC, y = Var)) +
  geom_bar(stat = "identity") +
  ggtitle("Scree Plot: Variance Explained by PCA Components") +
  ylab("Proportion of Variance Explained")

# PCA scatter plot: PC1 vs PC2
pca_df <- data.frame(
  Sample = rownames(t(expr_z_top)),
  PC1 = pca_res$x[,1],
  PC2 = pca_res$x[,2]
)

ggplot(pca_df, aes(x = PC1, y = PC2)) +
  geom_point(alpha = 0.7) +
  ggtitle("PCA of KIRC Patients (Top 3000 Variance Genes)") +
  xlab(paste0("PC1 (", round(pve[1] * 100, 1), "%)")) +
  ylab(paste0("PC2 (", round(pve[2] * 100, 1), "%)")) +
  theme_minimal()
```
# Select PCs explaining 80% variance
```{r}
# Compute cumulative variance explained
cumvar <- cumsum(pve)

# Find how many PCs are needed to reach ≥80% variance
Top_pc <- which(cumvar >= 0.80)[1]
cat("Number of PCs explaining ≥80% variance:", k, "\n")
```

# Extract the PC score 
```{r}
# PCA scores (samples × PCs)
pc_scores <- pca_res$x[, 1:Top_pc]
```


# Cluster on the main PCs
```{r}
# Compute distance on reduced PC space
dist_mat <- dist(pc_scores)

# Ward's hierarchical clustering
hc <- hclust(dist_mat, method = "ward.D2")

# Plot dendrogram
plot(hc, labels = FALSE, main = "Hierarchical Clustering on PCs (80% variance)")
rect.hclust(hc, k = 6, border = 2)
```

#Cutting Dendogram and 
We decided to cut the dendrogram at 6 since there is 6 subtypes of KIRC

```{r}
clusters <- cutree(hc, k = 6)
table(clusters)
```

# ============================================================================
# INVESTIGATION 2: Survival across the patient clusters is different. 
# RQ 2.2: Do expression-based clusters differ in patient outcomes?
# ============================================================================

# Create a dataframe that contains the DSS and the cluster labels. Rowname is patient ID
```{r}
# Create a dataframe that contains the cluster and the patient ID
clustering_data <- data.frame(
  clusters = clusters,
  PATIENT_ID = names(clusters)
)

# transform clustering data into standard patient ID
clustering_data$PATIENT_ID <- gsub("\\.", "-", substr(clustering_data$PATIENT_ID, 1, 12))

rownames(clustering_data) <- NULL

clinical_data_subset <- data_clinical_patient[
 data_clinical_patient$PATIENT_ID %in% clustering_data$PATIENT_ID,
 c("PATIENT_ID", "DSS_STATUS", "DSS_MONTHS")
]

# transform the DSS status to binary
clinical_data_subset$DSS_STATUS <- substr(clinical_data_subset$DSS_STATUS, 1, 1)

merged_data <- merge(
 clustering_data,
 clinical_data_subset,
 by = "PATIENT_ID",
 all.x = FALSE,
 all.y = FALSE
)
```


# Survival Analysis Across different clusters
```{r}
surv_obj_ward_DSS <- Surv(time = as.numeric(merged_data$DSS_MONTHS), event= as.numeric(merged_data$DSS_STATUS))

# KM plot for ward.D2 clustering
ward_d2_DSS <- surv_fit(surv_obj_ward_DSS ~ clusters, data = merged_data)
ward_d2_DSS_plot <- ggsurvplot(ward_d2_DSS,
           data = merged_data,
           pval = TRUE,
           conf.int = TRUE,
           title = "Disease Specific Survival for Identified Ward.D2 Hierarchical Clusters",
           xlab = "Time (Months)",
           ylab = "Survival Probability",
           ggtheme = theme_minimal(base_size = 14))

ward_d2_DSS_plot
```

# ============================================================================
# INVESTIGATION 3: Check DE across clusters
# RQ 2.2: What gene vary across clusters? Are they Significative of KIRC subtype?
# ============================================================================

#

# ============================================================================
# INVESTIGATION 4: 
# RQ 2.4:
# ============================================================================
```{r}
## Get Initial Count Matrix
# Create count matrix (genes as rows, samples as columns)
count_matrix <- as.matrix(RNAseq_KIRC[, -1])  # Remove gene name column
# Assuming your Ensembl IDs are in the first column of RNAseq_KIRC
ensemble_ids <- RNAseq_KIRC[, 1]
# Fix common Ensembl ID format issue (remove version number like ".1")
# This ensures a better match rate with the annotation package
ensemble_ids_clean <- sub("\\.\\d+$", "", ensemble_ids)
rownames(count_matrix) <- ensemble_ids_clean

cat("Initial count matrix dimensions (by Ensembl ID):", dim(count_matrix), "\n")

# Map Ensembl IDs to Gene Symbols
gene_map <- select(
  org.Hs.eg.db,
  keys = rownames(count_matrix),
  columns = c("SYMBOL"),
  keytype = "ENSEMBL"
)

# Filter out Ensembl IDs that didn't map to a Gene Symbol
gene_map <- gene_map %>% filter(!is.na(SYMBOL))

# Handle one-to-many mappings (Multiple Ensembl IDs map to one Symbol)
# For simplicity, we keep the first occurrence of a symbol
gene_map <- gene_map %>%
  distinct(ENSEMBL, .keep_all = TRUE) %>% # Keep one SYMBOL per ENSEMBL
  distinct(SYMBOL, .keep_all = TRUE) # Keep one row per unique SYMBOL

cat("Number of unique Ensembl IDs mapped to a unique Symbol:", nrow(gene_map), "\n")

# Filter and Rename the Count Matrix

# Filter the count matrix to include only mapped genes
mapped_count_matrix <- count_matrix[gene_map$ENSEMBL, ]

# Assign the Gene Symbols as row names
rownames(mapped_count_matrix) <- gene_map$SYMBOL

cat("Final count matrix dimensions (by Gene Symbol):", dim(mapped_count_matrix), "\n")
```


```{r}
# Recode age (make under80 the reference level)
age_info$age_group <- ifelse(age_info$age == "0-79", "under 80", "over 80")
age_info$age_group <- factor(age_info$age_group, levels = c("under 80", "over 80"))

```

```{r}
# Subset raw count matrix to matched samples
counts_sub <- mapped_count_matrix[, age_info$sample_id]
# Check dimensions
cat("Counts subset:", dim(counts_sub), "\n")
```

```{r}
# Filter: Keep genes with ≥10 counts in ≥20% of samples
min_samples <- 0.2 * ncol(counts_sub)   # ~85 samples

keep_genes <- rowSums(counts_sub >= 10) >= min_samples 
filtered_counts <- counts_sub[keep_genes, ]

cat("Filtered count matrix dimensions:", dim(filtered_counts), "\n")

```

```{r}
# Create DESeqDataset
dds <- DESeqDataSetFromMatrix(
    countData = filtered_counts,
    colData = data.frame(age_group = age_info$age_group, 
                         row.names = age_info$sample_id),
    design = ~ age_group
)
```

```{r}
# Run DESeq2
dds <- DESeq(dds)
res <- results(dds, contrast = c("age_group", "over 80", "under 80"))

```
```{r}
# Option 3: Super simple summary
if("ACSL6" %in% rownames(res)) {
  cat("ACSL6 Results:\n")
  cat("Log2 Fold Change:", round(res["ACSL6", "log2FoldChange"], 3), "\n")
  cat("P-value:", formatC(res["ACSL6", "pvalue"], format = "e", digits = 2), "\n")
  cat("Adjusted P-value:", formatC(res["ACSL6", "padj"], format = "e", digits = 2), "\n")
}
```

```{r}
if("ACSL6" %in% rownames(res)) {
  ACSL6_res <- res["ACSL6", ]
  print(ACSL6_res)
} else {
  print("ACSL6 not found in DESeq2 results — check symbol mapping.")
}
```
```{r}
# Get ACSL6 expression data
normalized_counts <- counts(dds, normalized = TRUE)
ACSL6_data <- data.frame(
  expression = normalized_counts["ACSL6", ],
  age_group = dds$age_group
)

# Simple boxplot
ggplot(ACSL6_data, aes(x = age_group, y = expression, fill = age_group)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +
  labs(title = "ACSL6 Expression by Age Group",
       x = "Age Group",
       y = "Expression") +
  theme_bw()

```
```{r}
ggplot(df_acsl6, aes(x = age_info$age_raw, y = expr)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "loess", se = TRUE, color = "blue") +
  labs(title = "ACSL6 Expression vs Age", x = "Age", y = "VST Expression") +
  theme_minimal()

```



####### Investigation 5

```{r}
# Filter out X and Y Chromosome Genes ---

# 1. Get chromosome information for the current gene set
# Note: Using Gene Symbols as keytype for a simpler mapping
gene_chr_map <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = rownames(mapped_count_matrix),
  columns = c("CHR"),
  keytype = "SYMBOL"
)

# 2. Identify Autosomal Genes (Chromosome 1-22)
# Exclude X, Y, MT (mitochondrial), and any unassigned/unknown chromosomes (NA)
# Autosomal chromosomes are '1' through '22'.
# We must first filter out NAs from the mapping step.
gene_chr_map <- gene_chr_map %>% filter(!is.na(CHR))

# Filter for only autosomes (1-22)
autosomal_gene_map <- gene_chr_map %>%
  filter(CHR %in% as.character(1:22)) %>%
  distinct(SYMBOL) # Keep only the unique autosomal symbols

# 3. Filter the count matrix to include only autosomal genes
autosomal_count_matrix <- mapped_count_matrix[autosomal_gene_map$SYMBOL, ]

cat("Number of genes removed (X, Y, MT, unassigned):",
    nrow(mapped_count_matrix) - nrow(autosomal_count_matrix), "\n")
cat("Count matrix dimensions (Autosomal Genes x Samples):",
    dim(autosomal_count_matrix), "\n")

# Reassign the count matrix variable to be used in the next step
count_data_to_filter <- autosomal_count_matrix
```


```{r}
# Ensure the sample order in the matrix and metadata match perfectly
# Filter the symbol-mapped counts to match samples in metadata
# Use the autosomal count matrix
count_data_final <- count_data_to_filter[, sample_metadata$sample_id]
col_data_final <- sample_metadata
...

cat("Count matrix dimensions (Genes x Samples):", dim(count_data_final), "\n")
cat("Metadata dimensions (Samples x Variables):", dim(col_data_final), "\n")
cat("Samples included in DESeq2 analysis:", nrow(col_data_final), "\n")
print(table(col_data_final$sex))
cat("\n")

# 1c. Create the DESeq2 object with the design formula: ~ sex
dds_sex <- DESeqDataSetFromMatrix(
  countData = count_data_final, # Now defined correctly
  colData = col_data_final,
  design = ~ sex
)

# Pre-filtering to remove genes with very low counts
# Remove more genes for efficiency and power
# Set the new minimum sample percentage (e.g., 50%)
min_percent_samples <- 0.50 
# Calculate the minimum number of samples required
min_samps <- ceiling(min_percent_samples * ncol(dds_sex))
# Filtering criteria: require >=10 counts in at least 'min_samps' samples
keep <- rowSums(counts(dds_sex) >= 10) >= min_samps
dds_sex <- dds_sex[keep,]
cat(sum(!keep), "genes removed due to low counts (using 50% sample cutoff).\n")
cat("Remaining genes for DESeq2 analysis:", nrow(dds_sex), "\n")
```

```{r}
# Run DESeq2
cat("Running DESeq analysis...\n")
dds_sex <- DESeq(dds_sex)
cat("DESeq analysis complete.\n\n")

# Extract the results: comparing the two levels of the 'sex' variable
# The comparison is 'Female' vs 'Male' if Female is the second factor level
results_sex <- results(dds_sex, name="sex_Male_vs_Female") # Adjust the 'name' based on your factor levels

# Filter and format results
results_df <- as.data.frame(results_sex) %>%
  rownames_to_column("gene") %>%
  filter(!is.na(padj)) %>% # Remove genes with NA p-values (often due to filtering or extreme values)
  arrange(padj)

# Get the count of significant genes
significant_genes_deseq <- sum(results_df$padj < 0.05)
cat("Significant genes (FDR < 0.05) via DESeq2:", significant_genes_deseq, "\n")
cat("Top 10 most differentially expressed genes by sex:\n")
print(head(results_df, 10))
```


```{r}
# Filter and format results
results_df <- as.data.frame(results_sex) %>%
  rownames_to_column("gene") %>%
  filter(!is.na(padj)) %>% # Remove genes with NA p-values
  
  # Sort by the absolute value of log2FoldChange for max variability
  arrange(desc(abs(log2FoldChange)))

# Get the count of significant genes
# Keep this part to assess DE significance
significant_genes_deseq <- sum(results_df$padj < 0.05)
cat("Significant autosomal genes (FDR < 0.05) via DESeq2:", significant_genes_deseq, "\n")
cat("Top 10 autosomal genes exhibiting the greatest variability in expression (by absolute log2FC):\n")
print(head(results_df, 10))
```
```{r}
# Filter for significant genes (FDR < 0.05) first, then sort by LFC magnitude
top_variability_genes <- results_df %>%
  filter(padj < 0.05) %>%
  # Arrange by the absolute log2FoldChange in descending order
  arrange(desc(abs(log2FoldChange)))

cat("Top 10 Autosomal Genes exhibiting the Greatest Variability (Highest |log2FC|):\n")
print(head(top_variability_genes, 10))
```
```{r}
# Get the list of significant Gene Symbols (your 2106 genes)
sig_genes_symbols <- results_df %>%
  filter(padj < 0.05) %>%
  pull(gene)

# Convert Gene Symbols to Entrez IDs
sig_genes_entrez <- bitr(
  sig_genes_symbols,
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Hs.eg.db
) %>%
  pull(ENTREZID)

cat("Number of unique significant autosomal genes successfully mapped to Entrez ID:", length(sig_genes_entrez), "\n")
```

```{r}
## GO Biological Process (BP) Enrichment
go_results <- enrichGO(
  gene          = sig_genes_entrez,
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENTREZID",
  ont           = "BP", # Biological Process
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05,
  readable      = TRUE
)

## KEGG Pathway Enrichment
kegg_results <- enrichKEGG(
  gene          = sig_genes_entrez,
  organism      = 'hsa',
  keyType       = "ncbi-geneid",  # make sure matches your IDs
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05,
  use_internal_data = FALSE        # uses internal annotation instead of online
)
kegg_results <- setReadable(kegg_results, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")


# Convert KEGG Entrez IDs to Gene Symbols for readability
kegg_results <- setReadable(kegg_results, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")

cat("\n--- Top 10 Enriched GO Biological Processes ---\n")
print(head(as.data.frame(go_results), 10))

cat("\n--- Top 10 Enriched KEGG Pathways ---\n")
print(head(as.data.frame(kegg_results), 10))
```

```{r}
# List genes contributing to top GO term
top_go_genes <- strsplit(go_results@result$geneID[1], "/")[[1]]
top_go_genes

```


```{r}
# 1. Select the top 20 most significant genes (by padj)
top_20_genes <- results_df %>%
  filter(padj < 0.05) %>%
  arrange(padj) %>%
  head(20) 

# 2. Get normalized, transformed counts from your DESeq2 object (dds_sex)
# Assuming dds_sex contains only autosomal genes after your filtering step
vsd <- vst(dds_sex, blind = FALSE)
vsd_matrix <- assay(vsd)

# 3. Filter the VSD matrix for the top genes and prepare annotation
top_20_vsd <- vsd_matrix[top_20_genes$gene, ]


# Create column annotation for sex
annotation_col <- as.data.frame(colData(dds_sex)["sex"])
rownames(annotation_col) <- rownames(colData(dds_sex))

# 4. Generate the Heatmap
pheatmap::pheatmap(
  top_20_vsd,
  cluster_rows = TRUE,
  show_rownames = TRUE,
  cluster_cols = TRUE,
  annotation_col = annotation_col,
  scale = "row",
  main = "Heatmap of Top 50 Autosomal Sex-Biased Genes",
  show_colnames = FALSE)
```


########Investigation 6

```{r}
# Subset to Samples(Patients) with Stage Info

# Extract patient IDs and match to clinical data
rnaseq_patient_ids <- substr(gsub("\\.", "-", colnames(expr_symbol)), 1, 12)

stage_info <- data.frame(
  sample_id = colnames(expr_symbol),
  patient_id = rnaseq_patient_ids
) %>%
  left_join(
    data_clinical_patient %>% 
      dplyr::select(PATIENT_ID, AJCC_PATHOLOGIC_TUMOR_STAGE) %>%
      rename(patient_id = PATIENT_ID, stage = AJCC_PATHOLOGIC_TUMOR_STAGE),
    by = "patient_id"
  ) %>%
  filter(!is.na(stage) & stage != "")

# Subset expression matrices
expr_staged <- expr_symbol[, stage_info$sample_id]
z_scores_staged <- z_scores_symbol[, stage_info$sample_id]

cat("Samples with stage:", nrow(stage_info), "\n")
print(table(stage_info$stage))
cat("\n")
```

```{r}
#Select top 500 Most Variable genes

# Calculate variance for each gene
gene_variance <- apply(expr_staged, 1, var)

# Select top 500
top500_genes <- names(sort(gene_variance, decreasing = TRUE)[1:500])

cat("Top 500 genes selected\n")
cat("Variance range:", round(range(gene_variance[top500_genes]), 2), "\n\n")

#This shows how expressed each gene is, and take the top 500 genes that have the largest variance in expression. 
#Each patient expresses a certain amount of each gene,
#so the highest variance is intersting as something might be happening, Could be indicative. 
```

```{r}
#ANOVA Test by stage

# Test if genes differ across stages
anova_results <- data.frame(
  gene = top500_genes,
  p_value = sapply(top500_genes, function(g) {
    fit <- aov(expr_staged[g, ] ~ factor(stage_info$stage))
    summary(fit)[[1]]["Pr(>F)"][1, 1]
  })
)
```

```{r}
#Multiple testing correction

anova_results$p_adj <- p.adjust(anova_results$p_value, method = "BH")
anova_results$significant <- anova_results$p_adj < 0.05

cat("Significant genes (FDR < 0.05):", sum(anova_results$significant), "\n")
print(anova_results %>% arrange(p_adj) %>% head(10))
cat("\n")
```

```{r}
#Heatmap Vizualization
#Out of the 500 most variable, only 85 are significant -> therefore, should only see small clustering....

top500_matrix <- z_scores_staged[top500_genes, stage_info$sample_id]

annotation <- data.frame(
  Stage = factor(stage_info$stage),
  row.names = stage_info$sample_id
)

pheatmap(
  top500_matrix,
  annotation_col = annotation,
  show_rownames = FALSE,
  show_colnames = FALSE,
  main = "Top 500 Variable Genes by Stage"
)
```

```{r}
#Identify the oncogenes

oncogenes <- c("MYC","KRAS","NRAS","HRAS","EGFR","ERBB2","PIK3CA",
               "BRAF","AKT1","AKT2","MDM2","CCND1","CDK4","CDK6",
               "MET","RET","ALK","ROS1","FGFR1","FGFR2")

oncogenes_in_data <- intersect(rownames(expr_staged), oncogenes)

cat("Oncogenes found:", length(oncogenes_in_data), "\n")
cat(paste(oncogenes_in_data, collapse = ", "), "\n\n")
```

```{r}
#Oncogene stage correlation
#Spearman correlation is a monotonic relationship between 2 variables
#In this case, we are looking at does expression go up or down with stage

if (length(oncogenes_in_data) > 0) {
  
  # Convert stage to numeric (I=1, II=2, III=3, IV=4)
  stage_numeric <- as.numeric(factor(
    stage_info$stage, 
    levels = c("STAGE I", "STAGE II", "STAGE III", "STAGE IV")
  ))
  
  # Test correlation for each oncogene
  oncogene_results <- data.frame(
    gene = oncogenes_in_data,
    correlation = sapply(oncogenes_in_data, function(g) {
      cor.test(stage_numeric, expr_staged[g, ], method = "spearman")$estimate
    }),
    p_value = sapply(oncogenes_in_data, function(g) {
      cor.test(stage_numeric, expr_staged[g, ], method = "spearman")$p.value
    })
  )
  
  oncogene_results$p_adj <- p.adjust(oncogene_results$p_value, method = "BH")
  oncogene_results$significant <- oncogene_results$p_adj < 0.05
  oncogene_results$direction <- ifelse(oncogene_results$correlation > 0, "UP", "DOWN")
  
  cat("Oncogene correlation results:\n")
  print(oncogene_results %>% arrange(p_adj))
  cat("\n")
}
```
```{r}
#Plot significant outcomes

if (exists("oncogene_results")) {
  sig_oncogenes <- oncogene_results %>% 
    filter(significant) %>% 
    pull(gene)
  
  if (length(sig_oncogenes) > 0) {
    for (gene in sig_oncogenes) {
      corr <- oncogene_results$correlation[oncogene_results$gene == gene]
      p_adj <- oncogene_results$p_adj[oncogene_results$gene == gene]
      
      plot_data <- data.frame(
        expression = expr_staged[gene, ],
        stage = factor(stage_info$stage)
      )
      
      p <- ggplot(plot_data, aes(stage, expression, fill = stage)) +
        geom_boxplot() +
        labs(
          title = paste(gene, "| ρ =", round(corr, 3), 
                       "| p_adj =", format(p_adj, scientific = TRUE, digits = 2)),
          x = "Tumor Stage", 
          y = "Log2 Expression"
        ) +
        theme_bw() +
        theme(legend.position = "none")
      
      print(p)
    }
  } else {
    cat("No significant oncogenes to plot\n")
  }
}
```

