---
title: "Final Project Part 2"
author:
- "Maximilien Filion - 20598645, "
- "Aneesh Kaura - 66832304, "
- "Monica Mihailescu - 47409289 "
date: "2025-11-17"
output:
  pdf_document: default
  html_document: default
  word_document: default
---
# ============================================================================
# RNA-seq Analysis Template: Clustering and Differential Expression
# ============================================================================
# This template provides a structured framework for comprehensive gene 
# expression analysis from preprocessing through clinical validation
# ============================================================================

# Load required libraries
```{r, results='hide'}
#if (!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install("org.Hs.eg.db")
# BiocManager::install("AnnotationDbi") 

# Install clusterProfiler properly
#BiocManager::install("clusterProfiler", ask = FALSE, update = FALSE)

#BiocManager::install("DOSE")
#BiocManager::install("pathview") 

library(clusterProfiler)
library(AnnotationDbi)
library(DESeq2)
library(pheatmap)
library(ggplot2)
library(cluster)
library(survival)
library(survminer)
#library(clusterProfiler)
library(org.Hs.eg.db) # Contains human gene annotations
library(dplyr)
library(tibble)
library(biomaRt)   
```


set.seed(1)

# ============================================================================
# STEP 1: DATA LOADING AND INITIAL SETUP
# ============================================================================

# Load the data files
# - count_matrix: raw counts (genes x samples)
# - clinical_data: patient clinical information
# - mutation_data: mutation status for each patient
```{r}
#Load the data (from lab 7)
RNAseq_KIRC <- read.csv("RNAseq_KIRC.csv")
data_clinical_patient <- read.delim("data_clinical_patient.txt", header = TRUE, sep = "\t", skip = 4)
data_mutations <- read.delim("data_mutations.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE)
```

```{r}
head(RNAseq_KIRC)
head(data_clinical_patient)
```


# Creates a dataframe containing all the patients common to the three datasets.
will be used in downstream applications.
```{r}
# RNAseq_KIRC dataset
rna_patient_ids <- gsub("\\.", "-", substr(colnames(RNAseq_KIRC), 1, 12))
# Data_clinical_patient dataset
clinical_patient_ids <- unique(data_clinical_patient$PATIENT_ID)
# Data_mutation datasetd
mutation_patient_ids <- unique(substr(data_mutations$Tumor_Sample_Barcode, 1, 12))
# Common patients
common_patients <- Reduce(intersect, list(rna_patient_ids,
clinical_patient_ids, mutation_patient_ids))

# Patients in all three datasets
common_patients <- Reduce(intersect, list(
  rna_patient_ids,
  clinical_patient_ids,
  mutation_patient_ids
))
cat("Number of patients in ALL datasets:", length(common_patients), "\n")
```

# Trim the datasets to only include the common patients
```{r}
#For the RNA seq dataset, we want to keep all the common patient columns and the ensembl gene ID (i.e.Column "X")
# Trim the RNA sequencing dataset
common_patients_and_X <- c(common_patients, "X")
RNAseq_KIRC <- RNAseq_KIRC[ , gsub("\\.", "-", substr(colnames(RNAseq_KIRC), 1, 12)) %in% common_patients_and_X]

# Trim the clinical dataset
data_clinical_patient <- data_clinical_patient[data_clinical_patient$PATIENT_ID %in% common_patients, ]

# Trim the mutation dataset
data_mutations <- data_mutations[substr(data_mutations$Tumor_Sample_Barcode, 1, 12) %in% common_patients, ]
```


<!-- # Merge the different ensembl version in the RNAseq dataframe -->
<!-- ```{r} -->
<!-- # Strip the ensembl version -->
<!-- RNAseq_KIRC$X <- gsub("\\..*$", "", RNAseq_KIRC$X) -->

<!-- # Collapse the duplicates -- GenAI was used for this part -->
<!-- RNAseq_KIRC_novers <- RNAseq_KIRC %>% -->
<!--   group_by(X) %>% -->
<!--   summarise(across(everything(), sum)) -->
<!-- ``` -->
The code above only cuts ~50 duplicate genes and takes a while to run. It was commented to signify that we thought about it but did not use it due to computational ressources. 


# Check for missing values
```{r}
# RNA-seq:
rnaseq_missing <- sum(is.na(RNAseq_KIRC))
cat("Missing values in RNA-seq count matrix:", rnaseq_missing, "\n")

# Clinical: Check key columns
key_clinical_cols <- c("PATIENT_ID", "AGE", "SEX", "OS_STATUS", "OS_MONTHS", "DSS_STATUS", "DSS_MONTHS", "DFS_STATUS", "DFS_MONTHS", "PFS_STATUS", "PFS_MONTHS", "PATH_T_STAGE", "PATH_N_STAGE", "PATH_M_STAGE", "AJCC_PATHOLOGIC_TUMOR_STAGE")
existing_cols <- key_clinical_cols[key_clinical_cols %in% colnames(data_clinical_patient)]

if(length(existing_cols) > 0) {
  for(col in existing_cols) {
    missing_count <- sum(is.na(data_clinical_patient[[col]]) | data_clinical_patient[[col]] == "")
    cat("Missing/empty values in", col, ":", missing_count, "\n")
  }
}

# Mutations: Check key columns
cat("\nMissing values in Tumor_Sample_Barcode:", sum(is.na(data_mutations$Tumor_Sample_Barcode)), "\n")
cat("Missing/empty values in Hugo_Symbol:", sum(is.na(data_mutations$Hugo_Symbol) | data_mutations$Hugo_Symbol == ""), "\n")
cat("Missing/empty values in Variant_Classificatoin:", sum(is.na(data_mutations$Variant_Classification) | data_mutations$Variant_Classificationl == ""), "\n")
cat("Missing/empty values in Consequence:", sum(is.na(data_mutations$Consequence) | data_mutations$Consequence == ""), "\n")
cat("Missing/epmty values in Impact:", sum(is.na(data_mutations$IMPACT) | data_mutations$IMPACT == ""), "\n\n")
```


# ============================================================================
# STEP 2: PREPROCESSING
# ============================================================================

## 2.1: Normalization using DESeq2
# GOAL: Account for sequencing depth and composition bias

# TODO: Create DESeq2 object from raw counts
# TODO: Estimate size factors
# TODO: Extract normalized counts
```{r}
# Create DESeq2 object from raw counts
# Prepare count matrix (genes as rows, samples as columns)
count_matrix <- as.matrix(RNAseq_KIRC[, -1])  # Remove gene name column
rownames(count_matrix) <- RNAseq_KIRC[, 1]    # Set gene names as row names

# Create a simple sample metadata data frame
# For normalization, we just need a basic colData with sample names
coldata <- data.frame(
  sample = colnames(count_matrix),
  condition = rep("tumor", ncol(count_matrix))  # Placeholder condition
)
rownames(coldata) <- coldata$sample

cat("Count matrix dimensions:", dim(count_matrix), "\n")

# Create DESeq2 dataset object
dds <- DESeqDataSetFromMatrix(
  countData = count_matrix,
  colData = coldata,
  design = ~ 1  # No DE, normalization only
)
cat("DESeq2 object created successfully\n\n")

# Estimate size factors
dds <- estimateSizeFactors(dds)

# Get size factors
size_factors <- sizeFactors(dds)
cat("Size factor summary:\n")
print(summary(size_factors))
cat("\n")

# Extract normalized counts
normalized_counts <- counts(dds, normalized = TRUE)
cat("Normalized count matrix dimensions:", dim(normalized_counts), "\n")
```

## 2.2: Log-transformation
# GOAL: Stabilize variance and make data more normally distributed

# TODO: Apply log2 transformation and VST
```{r}
# simple Log2(x+1) transformation 
log_norm_counts <- log2(normalized_counts + 1)

cat("Log transform count matrix dimensions:", dim(log_norm_counts), "\n")
cat("Log transformation range:", range(log_norm_counts), "\n\n")

# Apply VST (recommended for >30 samples) - Applies variance-stabilizing transformation, 'blind' Ignores experimental design
vsd <- vst(dds, blind = TRUE)
vst_counts <- assay(vsd)

cat("VST count matrix dimensions:", dim(vst_counts), "\n")
cat("VST value range:", range(vst_counts), "\n\n")
```

## 2.3: Z-score transformation
# GOAL: Standardize gene expression across samples

# TODO: Calculate z-scores per gene (across samples)
# TODO: Verify mean ≈ 0 and sd ≈ 1 for each gene
```{r}
# Z-score the log-transformed counts (by gene)
z_log <- t(scale(t(log_norm_counts)))

cat("Z-scored log2 matrix dimensions:", dim(z_log), "\n")
cat("Z-scored log2 range:", range(z_log, na.rm = TRUE), "\n\n")

# Z-score the VST-transformed counts (by gene)
z_vst <- t(scale(t(vst_counts)))

cat("Z-scored VST matrix dimensions:", dim(z_vst), "\n")
cat("Z-scored VST range:", range(z_vst, na.rm = TRUE), "\n\n")
```


# ============================================================================
# INVESTIGATION 1: Global Expression-Based Clustering
# RQ 2.1: Do patients cluster into distinct phenotype subtypes?
# ============================================================================

# Filter the most variable genes 
Running distance on 60660 observations was too computationally intensive. We decided to only keep the top 3000 genes since PCA will explain only part of the variance and variance will be mostly explained by those highly variable 3000 genes. It was chosen arbitrarily 

```{r}
# compute variance on raw data
gene_vars <- apply(normalized_counts, 1, var)


#Rank the genes by variance
top_genes <- names(sort(gene_vars, decreasing = TRUE))[1:3000]

# Create a subset matrix with the top 3000 genes
expr_z_top <- z_vst[top_genes, ]
```


# Plot PCA and look for distinct group
```{r}
pca_res <- prcomp(t(expr_z_top), scale = TRUE)

# Percentage variance explained
pve <- (pca_res$sdev^2) / sum(pca_res$sdev^2)

# Create and save Scree plot
scree_plot <- ggplot(data.frame(PC = 1:10, Var = pve[1:10]),
       aes(x = PC, y = Var)) +
  geom_bar(stat = "identity") +
  ggtitle("Scree Plot: Variance Explained by PCA Components") +
  ylab("Proportion of Variance Explained")

ggsave("scree_plot.png", scree_plot, width = 7, height = 5, dpi = 300)

# Create and save PCA scatter plot
pca_df <- data.frame(
  Sample = rownames(t(expr_z_top)),
  PC1 = pca_res$x[,1],
  PC2 = pca_res$x[,2]
)

pca_scatter <- ggplot(pca_df, aes(x = PC1, y = PC2)) +
  geom_point(alpha = 0.7) +
  ggtitle("PCA of KIRC Patients (Top 3000 Variance Genes)") +
  xlab(paste0("PC1 (", round(pve[1] * 100, 1), "%)")) +
  ylab(paste0("PC2 (", round(pve[2] * 100, 1), "%)")) +
  theme_minimal()

ggsave("pca_scatter.png", pca_scatter, width = 7, height = 6, dpi = 300)
```
# Select PCs explaining 80% variance
```{r}
# Compute cumulative variance explained
cumvar <- cumsum(pve)

# Find how many PCs are needed to reach ≥80% variance
Top_pc <- which(cumvar >= 0.80)[1]
cat("Number of PCs explaining ≥80% variance:", Top_pc, "\n")
```

# Extract the PC score 
```{r}
# PCA scores (samples × PCs)
pc_scores <- pca_res$x[, 1:Top_pc]
```


# Cluster on the main PCs
```{r}
# Compute distance on reduced PC space
dist_mat <- dist(pc_scores)

# Ward's hierarchical clustering
hc <- hclust(dist_mat, method = "ward.D2")

# Plot dendrogram
plot(hc, labels = FALSE, main = "Hierarchical Clustering on PCs (80% variance)")
rect.hclust(hc, k = 3, border = 2)

png("dendrogram.png", width = 2400, height = 1800, res = 300)
plot(hc, labels = FALSE, main = "Hierarchical Clustering on PCs (80% variance)")
rect.hclust(hc, k = 3, border = 2)
dev.off()
```

#Cutting Dendogram
We decided to cut the dendrogram at 6 since there is 6 subtypes of KIRC

```{r}
clusters <- cutree(hc, k = 3)
table(clusters)
```
# Plot Overlapping Elbow and silhouette graphs
Do not pick 2, Since it was proven online that they might be artifacts of batch effects
```{r}
# install.packages("factoextra")
library(factoextra)

# Silhouette width 
silhouette <- fviz_nbclust(as.matrix(dist_mat), hcut, method = "silhouette", hc_method = "ward.D2")
silhouette

# Elbow plot
withinsumsquares <- fviz_nbclust(as.matrix(dist_mat), hcut, method = "wss", hc_method = "ward.D2")
withinsumsquares

ggsave("silhouette_plot.png", silhouette, width = 7, height = 5, dpi = 300)
ggsave("elbow_plot.png", withinsumsquares, width = 7, height = 5, dpi = 300)
```


# ============================================================================
# INVESTIGATION 2: Survival across the patient clusters is different. 
# RQ 2.2: Do expression-based clusters differ in patient outcomes?
# ============================================================================

# Create a dataframe that contains the DSS and the cluster labels. Rowname is patient ID
```{r}
# Create a dataframe that contains the cluster and the patient ID
clustering_data <- data.frame(
  clusters = clusters,
  PATIENT_ID = names(clusters)
)

# transform clustering data into standard patient ID
clustering_data$PATIENT_ID <- gsub("\\.", "-", substr(clustering_data$PATIENT_ID, 1, 12))

rownames(clustering_data) <- NULL

clinical_data_subset <- data_clinical_patient[
 data_clinical_patient$PATIENT_ID %in% clustering_data$PATIENT_ID,
 c("PATIENT_ID", "DSS_STATUS", "DSS_MONTHS")
]

# transform the DSS status to binary
clinical_data_subset$DSS_STATUS <- substr(clinical_data_subset$DSS_STATUS, 1, 1)

merged_data <- merge(
 clustering_data,
 clinical_data_subset,
 by = "PATIENT_ID",
 all.x = FALSE,
 all.y = FALSE
)
```


# Survival Analysis Across different clusters
```{r}
surv_obj_ward_DSS <- Surv(time = as.numeric(merged_data$DSS_MONTHS), event= as.numeric(merged_data$DSS_STATUS))

# KM plot for ward.D2 clustering
ward_d2_DSS <- surv_fit(surv_obj_ward_DSS ~ clusters, data = merged_data)
ward_d2_DSS_plot <- ggsurvplot(ward_d2_DSS,
           data = merged_data,
           pval = TRUE,
           conf.int = TRUE,
           title = "Disease Specific Survival for Identified Ward.D2 Hierarchical Clusters",
           xlab = "Time (Months)",
           ylab = "Survival Probability",
           ggtheme = theme_minimal(base_size = 14))

ward_d2_DSS_plot

ggsave("DSSacrossclusters.png", ward_d2_DSS_plot$plot, width=7, height=3, dpi=300)
```
Since we see 3 clear outcomes in survival. This is good since there is 3 histological subtypes

# Assign the clusters to respective patients and their data
```{r}
# Create a dataframe that contains the cluster and the patient ID
clustering_data <- data.frame(
  clusters = clusters,
  PATIENT_ID = names(clusters)
)

# Transform clustering data into standard patient ID
clustering_data$PATIENT_ID <- gsub("\\.", "-", substr(clustering_data$PATIENT_ID, 1, 12))
rownames(clustering_data) <- NULL

# Extract relevant clinical data
clinical_data_subset <- data_clinical_patient[
  data_clinical_patient$PATIENT_ID %in% clustering_data$PATIENT_ID,
  c("PATIENT_ID", "DSS_STATUS", "DSS_MONTHS", "AGE", "AJCC_PATHOLOGIC_TUMOR_STAGE")
]

# Transform the DSS status to binary (0 = alive, 1 = dead)
clinical_data_subset$DSS_STATUS <- substr(clinical_data_subset$DSS_STATUS, 1, 1)

# Transform the tumor stage
clinical_data_subset$AJCC_PATHOLOGIC_TUMOR_STAGE <- sapply(clinical_data_subset$AJCC_PATHOLOGIC_TUMOR_STAGE, function(x) {
  if (grepl("stage", x, ignore.case = TRUE)) {
            as.numeric(as.roman(gsub("stage ", "", x, ignore.case = TRUE)))
  } else {
    NA  
  }
})

# Merge clustering and clinical data
merged_data <- merge(
  clustering_data,
  clinical_data_subset,
  by = "PATIENT_ID",
  all.x = FALSE,
  all.y = FALSE
)
```


# Look for diffences in age - ANOVA. Code adapted from part 1 of the project.
```{r}
# One-way ANOVA for age
merged_data$clusters <- as.factor(merged_data$clusters)

anova_result_Ward_D2 <- aov(merged_data$AGE ~ clusters, data = merged_data)
summary(anova_result_Ward_D2)

TukeyHSD(anova_result_Ward_D2)

anova_ward_d2_plot <- ggplot(merged_data, aes(x = clusters, y = AGE, fill = clusters)) +
  geom_boxplot() +
  labs(
    title = "Patient Age Distribution Across Ward.D2 Clusters",
    x = "Cluster",
    y = "Age (years)"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")

# Plot
anova_ward_d2_plot

ggsave("AgeDistributionAcrossClusters_WardD2.png", anova_ward_d2_plot, width=7, height=3, dpi=300)
```

# What about Tumor stage?
```{r}
merged_data$AJCC_PATHOLOGIC_TUMOR_STAGE <- as.factor(merged_data$AJCC_PATHOLOGIC_TUMOR_STAGE)

# Contingency table
table_stage_cluster_wardd2 <- table(merged_data$clusters, merged_data$AJCC_PATHOLOGIC_TUMOR_STAGE)
table_stage_cluster_wardd2

# Chi-square Test
chisq_test_wardd2 <- chisq.test(table_stage_cluster_wardd2)
chisq_test_wardd2

# Visualization
chi_test_plot_ward2 <- ggplot(merged_data, aes(x = clusters, fill = AJCC_PATHOLOGIC_TUMOR_STAGE)) +
  geom_bar(position = "fill") +  # fill = proportion per cluster
  coord_flip() +
  ylab("Proportion") +
  xlab("Cluster") +
  labs(title = "Tumor Stage Across Ward.D2 Clusters", fill = "Tumor Stage") +
  theme_minimal(base_size = 14)

#plot
chi_test_plot_ward2

#save
ggsave("TumorStageAcrossClusters_WardD2.png",chi_test_plot_ward2 , width=7, height=3, dpi=300)
```




# ============================================================================
# INVESTIGATION 3: Check DE across clusters
# RQ 2.2: What gene vary across clusters? Are they Significative of KIRC subtype?
# ============================================================================

# ============================================================================
# Run DESeq2 for Differential Expression Between Clusters
# ============================================================================
This step was made by claudeAI usign the prompt(s):"Read this + Previous code, and make a DE heatmap using pheatmap" 
```{r}
cat("Merged data dimensions:", dim(merged_data), "\n")
cat("Cluster distribution:\n")
print(table(merged_data$clusters))
cat("\n")

# ============================================================================
# Run DESeq2 for Differential Expression Between Clusters
# ============================================================================

# Prepare DESeq2 object with cluster information
# Use the original sample names (with dots/full barcodes)
coldata_clusters <- data.frame(
  sample = colnames(count_matrix),
  cluster = clusters[colnames(count_matrix)]
)
rownames(coldata_clusters) <- coldata_clusters$sample
coldata_clusters$cluster <- as.factor(coldata_clusters$cluster)

# Create DESeq2 dataset with cluster design
dds_de <- DESeqDataSetFromMatrix(
  countData = count_matrix,
  colData = coldata_clusters,
  design = ~ cluster
)

# Filter low-count genes (at least 10 reads in at least 10% of samples)
keep <- rowSums(counts(dds_de) >= 10) >= ncol(dds_de) * 0.1
dds_de <- dds_de[keep, ]
cat("Genes retained after filtering:", nrow(dds_de), "\n\n")

# Run DESeq2
dds_de <- DESeq(dds_de)

# ============================================================================
# Extract Top Differentially Expressed Genes
# ============================================================================

# Get all pairwise comparisons
comparisons <- list(
  c1_vs_c2 = results(dds_de, contrast = c("cluster", "1", "2")),
  c1_vs_c3 = results(dds_de, contrast = c("cluster", "1", "3")),
  c2_vs_c3 = results(dds_de, contrast = c("cluster", "2", "3"))
)

# Find significant DEGs for each comparison (padj < 0.05, |log2FC| > 1)
sig_genes_list <- lapply(comparisons, function(res) {
  sig <- res[which(res$padj < 0.05 & abs(res$log2FoldChange) > 1), ]
  rownames(sig)
})

# Combine all significant genes
all_sig_genes <- unique(unlist(sig_genes_list))
cat("Total unique DEGs across all comparisons:", length(all_sig_genes), "\n")

# Select top genes by variance if there are too many
if (length(all_sig_genes) > 500) {
  gene_vars_sig <- apply(vst_counts[all_sig_genes, ], 1, var)
  top_sig_genes <- names(sort(gene_vars_sig, decreasing = TRUE))[1:500]
  cat("Selected top 500 DEGs by variance\n\n")
} else {
  top_sig_genes <- all_sig_genes
}
```
# Create Heatmap
```{r}
# Extract VST counts for top DEGs
heatmap_data <- vst_counts[top_sig_genes, ]

# Z-score normalize by gene (rows)
heatmap_data_scaled <- t(scale(t(heatmap_data)))

# Order samples by cluster
sample_order <- order(clusters)
heatmap_data_scaled <- heatmap_data_scaled[, sample_order]
clusters_ordered <- clusters[sample_order]

# Create annotation for samples (columns)
annotation_col <- data.frame(
  Cluster = as.factor(clusters_ordered)
)
rownames(annotation_col) <- colnames(heatmap_data_scaled)

# Define colors for clusters
cluster_colors <- list(
  Cluster = c("1" = "#E69F00", "2" = "#56B4E9", "3" = "#009E73")
)
```


# Save the image
```{r}
# Create the heatmap
DEheatmap <- pheatmap(
  heatmap_data_scaled,
  scale = "none",  # Already scaled
  cluster_rows = TRUE,  # Cluster genes
  cluster_cols = FALSE,  # Keep sample order by cluster
  clustering_distance_rows = "euclidean",
  clustering_method = "ward.D2",
  annotation_col = annotation_col,
  annotation_colors = cluster_colors,
  show_colnames = FALSE,  # Too many samples
  show_rownames = FALSE,  # Too many genes
  color = colorRampPalette(c("green", "black", "red"))(100),
  main = "Differential Gene Expression Across Clusters",
  fontsize = 10,
  border_color = NA,
  #filename = "DEpercluster.png"  # Add this parameter
)
```
# Which genes are the most differentially expressed? 
Code belwo was generated with ClaudeAI and the prompts: I want as simple as possible, in R so I can see hte comparisons. + Code aboe
```{r}
library(ggplot2)
library(gridExtra)

# Function to create volcano plot
make_volcano <- function(res, title) {
  # Prepare data
  df <- as.data.frame(res)
  df$gene <- rownames(df)
  df$neglog10padj <- -log10(df$padj)
  
  # Add significance category
  df$significance <- "Not Significant"
  df$significance[df$padj < 0.05 & df$log2FoldChange > 1] <- "Upregulated"
  df$significance[df$padj < 0.05 & df$log2FoldChange < -1] <- "Downregulated"
  df$significance <- factor(df$significance, 
                           levels = c("Upregulated", "Downregulated", "Not Significant"))
  
  # Remove NA values
  df <- df[!is.na(df$padj) & !is.na(df$log2FoldChange), ]
  
  # Count genes in each category
  n_up <- sum(df$significance == "Upregulated")
  n_down <- sum(df$significance == "Downregulated")
  
  # Create plot
  p <- ggplot(df, aes(x = log2FoldChange, y = neglog10padj, color = significance)) +
    geom_point(alpha = 0.6, size = 1.5) +
    scale_color_manual(values = c("Upregulated" = "#ef4444", 
                                  "Downregulated" = "#3b82f6", 
                                  "Not Significant" = "#9ca3af")) +
    geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "gray50") +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray50") +
    labs(title = title,
         subtitle = paste0("Up: ", n_up, " | Down: ", n_down),
         x = "log2 Fold Change",
         y = "-log10(adjusted p-value)",
         color = "Significance") +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          plot.subtitle = element_text(hjust = 0.5),
          legend.position = "bottom")
  
  return(p)
}

# Create volcano plots for each comparison
p1 <- make_volcano(comparisons$c1_vs_c2, "Cluster 1 vs Cluster 2")
p2 <- make_volcano(comparisons$c1_vs_c3, "Cluster 1 vs Cluster 3")
p3 <- make_volcano(comparisons$c2_vs_c3, "Cluster 2 vs Cluster 3")

# Display all three plots
grid.arrange(p1, p2, p3, ncol = 3)

# Optional: Save as high-resolution figure
ggsave("volcano_c1_vs_c2.png", p1, width = 8, height = 6, dpi = 300)
ggsave("volcano_c1_vs_c3.png", p2, width = 8, height = 6, dpi = 300)
ggsave("volcano_c2_vs_c3.png", p3, width = 8, height = 6, dpi = 300)
```

# Which genes (by name) are the most differentially expressed?
```{r}
cat("\n=== Differential Expression Summary ===\n")
for(comp_name in names(comparisons)) {
  res <- comparisons[[comp_name]]
  df <- as.data.frame(res)
  df$gene <- rownames(df)
  df <- df[!is.na(df$padj), ]
  
  n_up <- sum(df$padj < 0.05 & df$log2FoldChange > 1, na.rm = TRUE)
  n_down <- sum(df$padj < 0.05 & df$log2FoldChange < -1, na.rm = TRUE)
  n_total_sig <- sum(df$padj < 0.05, na.rm = TRUE)
  
  cat(sprintf("\n%s:\n", comp_name))
  cat(sprintf("  Upregulated (padj<0.05, log2FC>1): %d\n", n_up))
  cat(sprintf("  Downregulated (padj<0.05, log2FC<-1): %d\n", n_down))
  cat(sprintf("  Total significant (padj<0.05): %d\n", n_total_sig))
  
  # Get top upregulated genes
  cat("\n  Top 10 UPREGULATED genes:\n")
  top_up <- df[df$padj < 0.05 & df$log2FoldChange > 0, ]
  top_up <- top_up[order(top_up$log2FoldChange, decreasing = TRUE), ]
  if(nrow(top_up) > 0) {
    for(i in 1:min(10, nrow(top_up))) {
      cat(sprintf("    %2d. %-15s | log2FC: %6.2f | padj: %.2e\n", 
                  i, top_up$gene[i], top_up$log2FoldChange[i], top_up$padj[i]))
    }
  } else {
    cat("    No significantly upregulated genes\n")
  }
  
  # Get top downregulated genes
  cat("\n  Top 10 DOWNREGULATED genes:\n")
  top_down <- df[df$padj < 0.05 & df$log2FoldChange < 0, ]
  top_down <- top_down[order(top_down$log2FoldChange, decreasing = FALSE), ]
  if(nrow(top_down) > 0) {
    for(i in 1:min(10, nrow(top_down))) {
      cat(sprintf("    %2d. %-15s | log2FC: %6.2f | padj: %.2e\n", 
                  i, top_down$gene[i], top_down$log2FoldChange[i], top_down$padj[i]))
    }
  } else {
    cat("    No significantly downregulated genes\n")
  }
  cat("\n", strrep("-", 70), "\n")
}
```




# ============================================================================
# INVESTIGATION 4: 
# RQ 2.4:
# ============================================================================
```{r}
# Extract sample IDs from your count matrix column names
sample_ids <- colnames(mapped_count_matrix)

# Clean up sample IDs to match patient IDs (remove any extra characters)
patient_ids <- gsub("\\.", "-", substr(sample_ids, 1, 12))

# Create age_info by matching with clinical data
age_info <- data.frame(
  sample_id = sample_ids,
  patient_id = patient_ids,
  stringsAsFactors = FALSE
)

# Merge with clinical data to get age information
age_info <- merge(
  age_info,
  data_clinical_patient[, c("PATIENT_ID", "AGE")], 
  by.x = "patient_id",
  by.y = "PATIENT_ID",
  all.x = TRUE
)

# Create age groups based on the AGE column
# Categorize into "0-79" and "80+" groups
age_info$age <- ifelse(age_info$AGE < 80, "0-79", "80+")

# Recode age (make under80 the reference level)
age_info$age_group <- ifelse(age_info$age == "0-79", "under 80", "over 80")
age_info$age_group <- factor(age_info$age_group, levels = c("under 80", "over 80"))

# Verify the structure
str(age_info)
head(age_info)
table(age_info$age_group)
```


```{r}
# Recode age (make under80 the reference level)
age_info$age_group <- ifelse(age_info$age == "0-79", "under 80", "over 80")
age_info$age_group <- factor(age_info$age_group, levels = c("under 80", "over 80"))

```

```{r}
# Subset raw count matrix to matched samples
counts_sub <- mapped_count_matrix[, age_info$sample_id]
# Check dimensions
cat("Counts subset:", dim(counts_sub), "\n")
```

```{r}
# Filter: Keep genes with ≥10 counts in ≥20% of samples
min_samples <- 0.2 * ncol(counts_sub)   # ~85 samples

keep_genes <- rowSums(counts_sub >= 10) >= min_samples 
filtered_counts <- counts_sub[keep_genes, ]

cat("Filtered count matrix dimensions:", dim(filtered_counts), "\n")

```

```{r}
# Create DESeqDataset
dds <- DESeqDataSetFromMatrix(
    countData = filtered_counts,
    colData = data.frame(age_group = age_info$age_group, 
                         row.names = age_info$sample_id),
    design = ~ age_group
)
```

```{r}
# Run DESeq2
dds <- DESeq(dds)
res <- results(dds, contrast = c("age_group", "over 80", "under 80"))

```

```{r}
# Super simple summary
if("ACSL6" %in% rownames(res)) {
  cat("ACSL6 Results:\n")
  cat("Log2 Fold Change:", round(res["ACSL6", "log2FoldChange"], 3), "\n")
  cat("P-value:", formatC(res["ACSL6", "pvalue"], format = "e", digits = 2), "\n")
  cat("Adjusted P-value:", formatC(res["ACSL6", "padj"], format = "e", digits = 2), "\n")
}
```

```{r}
if("ACSL6" %in% rownames(res)) {
  ACSL6_res <- res["ACSL6", ]
  print(ACSL6_res)
} else {
  print("ACSL6 not found in DESeq2 results — check symbol mapping.")
}
```


```{r}
# Get ACSL6 expression data
normalized_counts <- counts(dds, normalized = TRUE)
ACSL6_data <- data.frame(
  expression = normalized_counts["ACSL6", ],
  age_group = dds$age_group
)

# Simple boxplot
ggplot(ACSL6_data, aes(x = age_group, y = expression, fill = age_group)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +
  labs(title = "ACSL6 Expression by Age Group",
       x = "Age Group",
       y = "Expression") +
  theme_bw()

```


####### Investigation 5

```{r}
# Filter out X and Y Chromosome Genes 

# Get chromosome information for the current gene set
# Note: Using Gene Symbols as keytype for a simpler mapping
gene_chr_map <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = rownames(mapped_count_matrix),
  columns = c("CHR"),
  keytype = "SYMBOL"
)

# Identify Autosomal Genes (Chromosome 1-22)
# Exclude X, Y, MT (mitochondrial), and any unassigned/unknown chromosomes (NA)
# Autosomal chromosomes are '1' through '22'.
# We must first filter out NAs from the mapping step.
gene_chr_map <- gene_chr_map %>% filter(!is.na(CHR))

# Filter for only autosomes (1-22)
autosomal_gene_map <- gene_chr_map %>%
  filter(CHR %in% as.character(1:22)) %>%
  distinct(SYMBOL) # Keep only the unique autosomal symbols

# Filter the count matrix to include only autosomal genes
autosomal_count_matrix <- mapped_count_matrix[autosomal_gene_map$SYMBOL, ]

cat("Number of genes removed (X, Y, MT, unassigned):",
    nrow(mapped_count_matrix) - nrow(autosomal_count_matrix), "\n")
cat("Count matrix dimensions (Autosomal Genes x Samples):",
    dim(autosomal_count_matrix), "\n")

# Reassign the count matrix variable to be used in the next step
count_data_to_filter <- autosomal_count_matrix
```

```{r}
# Check that sex-chromosome linked genes were removed
# Get chromosome mapping for the filtered (autosomal) count matrix
filtered_chr_map <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = rownames(count_data_to_filter),
  columns = c("CHR"),
  keytype = "SYMBOL"
)

# Count how many genes map to X or Y
sum(filtered_chr_map$CHR %in% c("X", "Y"))

# List any genes on X or Y (should be empty)
filtered_chr_map %>% filter(CHR %in% c("X", "Y"))
# if output is empty then no x/y genes remain
```
```{r}
# Extract sample IDs from your count matrix column names
# Use the autosomal count matrix bc filtering sex chromosomes
sample_ids <- colnames(count_data_to_filter)

# Clean up sample IDs to match patient IDs
patient_ids <- gsub("\\.", "-", substr(sample_ids, 1, 12))

# Create sex_info by matching with clinical data
sex_info <- data.frame(
  sample_id = sample_ids,
  patient_id = patient_ids,
  stringsAsFactors = FALSE
)

# Merge with clinical data to get sex information
sex_info <- merge(
  sex_info,
  data_clinical_patient[, c("PATIENT_ID", "SEX")],  # Adjust "SEX" if column name differs
  by.x = "patient_id",
  by.y = "PATIENT_ID",
  all.x = TRUE
)

# Rename the SEX column to lowercase for consistency
names(sex_info)[names(sex_info) == "SEX"] <- "sex"

# Filter out any missing or empty sex values
sex_info <- sex_info %>%
  filter(!is.na(sex) & sex != "")

# Set sex as a factor with Male as reference level
sex_info$sex <- factor(sex_info$sex, levels = c("Male", "Female"))

# Verify the structure
str(sex_info)
head(sex_info)
table(sex_info$sex) 

# Check that sample_ids match between sex_info and count matrix
cat("Samples in sex_info:", nrow(sex_info), "\n")
cat("Samples in count matrix:", ncol(count_data_to_filter), "\n")
```


```{r}
# Subset the mapped count matrix to samples that have sex information
counts_sub_sex <- count_data_to_filter[, sex_info$sample_id]

# Check dimensions
cat("Counts subset for sex comparison:", dim(counts_sub_sex), "\n")

```

```{r}
# Apply low count filter
min_samples_sex <- 0.2 * ncol(counts_sub_sex)

keep_genes_sex <- rowSums(counts_sub_sex >= 10) >= min_samples_sex

filtered_counts_sex <- counts_sub_sex[keep_genes_sex, ]
sex_info$sex <- factor(sex_info$sex, levels = c("Male", "Female"))

```


```{r}
dds_sex <- DESeqDataSetFromMatrix(
    countData = filtered_counts_sex,
    colData = data.frame(
        sex = sex_info$sex,
        row.names = sex_info$sample_id
    ),
    design = ~ sex
)
```

```{r}
# Pre-filter low counts 
# Keep genes with at least 10 reads across all samples
dds_sex <- dds_sex[rowSums(counts(dds_sex)) >= 10, ]

# Run DESeq2 
dds_sex <- DESeq(dds_sex)
```

```{r}
colnames(colData(dds_sex))
```

```{r}
# Get results
res_sex <- results(dds_sex, contrast = c("sex", "Female", "Male")) 
# This gives log2FC = Female vs Male

# Sort by adjusted p-value
res_sex <- res_sex[order(res_sex$padj), ]

# View top genes
head(res_sex, 10)

# Optional: convert to a data frame for easier manipulation
res_df <- as.data.frame(res_sex)
res_df$gene <- rownames(res_df)

# Filter for significantly DE genes (FDR < 0.05)
sig_res <- res_df[!is.na(res_df$padj) & res_df$padj < 0.05, ]

# Order by absolute log2 fold change, largest first
sig_res_absFC <- sig_res[order(abs(sig_res$log2FoldChange), decreasing = TRUE), ]

#console shows top most significant genes
#df shows most differentially expressed genes (by log2fc)
# View top 10 genes with largest absolute log2FC
head(sig_res_absFC, 10)

```

```{r}
# Vector of genes to inspect
genes_of_interest <- c("CCDC146", "UGT1A6")

# Subset DESeq2 results for these genes
res_subset <- res_sex[rownames(res_sex) %in% genes_of_interest, ]

# If you want to see it as a clean data frame
res_subset_df <- as.data.frame(res_subset)
res_subset_df$gene <- rownames(res_subset_df)  # add gene column for clarity

# Print results
print(res_subset_df)
```


#Investigation 6

```{r}
library(ggplot2)
library(pheatmap)
library(biomaRt)  
library(dplyr)
```

```{r}
#Clean Ensembl IDs and map to gene symbols

# Remove version numbers from Ensembl IDs
clean_ids <- sub("\\..*", "", rownames(log_norm_counts))
rownames(log_norm_counts) <- clean_ids
rownames(z_log) <- clean_ids

# Get gene symbols from Ensembl
mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
mapping <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters = "ensembl_gene_id",
  values = clean_ids,
  mart = mart
)

# Convert to gene symbols
expr_df <- as.data.frame(log_norm_counts) %>%
  mutate(ensembl_gene_id = rownames(log_norm_counts)) %>%
  left_join(mapping, by = "ensembl_gene_id") %>%
  filter(hgnc_symbol != "") %>%
  distinct(hgnc_symbol, .keep_all = TRUE)

rownames(expr_df) <- expr_df$hgnc_symbol
expr_symbol <- as.matrix(expr_df[, -c(ncol(expr_df)-1, ncol(expr_df))])

# Same for z-scores
z_df <- as.data.frame(z_log) %>%
  mutate(ensembl_gene_id = rownames(z_log)) %>%
  left_join(mapping, by = "ensembl_gene_id") %>%
  filter(hgnc_symbol != "") %>%
  distinct(hgnc_symbol, .keep_all = TRUE)

rownames(z_df) <- z_df$hgnc_symbol
z_scores_symbol <- as.matrix(z_df[, -c(ncol(z_df)-1, ncol(z_df))])

cat("Genes with symbols:", nrow(expr_symbol), "\n\n")

```


```{r}
# Subset to Samples(Patients) with Stage Info

# Extract patient IDs and match to clinical data
rnaseq_patient_ids <- substr(gsub("\\.", "-", colnames(expr_symbol)), 1, 12)

stage_info <- data.frame(
  sample_id = colnames(expr_symbol),
  patient_id = rnaseq_patient_ids
) %>%
  left_join(
    data_clinical_patient %>%
      dplyr::select(PATIENT_ID, AJCC_PATHOLOGIC_TUMOR_STAGE) %>%
      rename(patient_id = PATIENT_ID, stage = AJCC_PATHOLOGIC_TUMOR_STAGE),
    by = "patient_id"
  ) %>%
  filter(!is.na(stage) & stage != "")

# Subset expression matrices
expr_staged <- expr_symbol[, stage_info$sample_id]
z_scores_staged <- z_scores_symbol[, stage_info$sample_id]

cat("Samples with stage:", nrow(stage_info), "\n")
print(table(stage_info$stage))
cat("\n")
```

```{r}
#Select top 500 Most Variable genes

# Calculate variance for each gene
gene_variance <- apply(expr_staged, 1, var)

# Select top 500
top500_genes <- names(sort(gene_variance, decreasing = TRUE)[1:500])

cat("Top 500 genes selected\n")
cat("Variance range:", round(range(gene_variance[top500_genes]), 2), "\n\n")

#This shows how expressed each gene is, and take the top 500 genes that have the largest variance in expression.
#Each patient expresses a certain amount of each gene,
#so the highest variance is intersting as something might be happening, Could be indicative.
```

```{r}
#ANOVA Test by stage

# Test if genes differ across stages
anova_results <- data.frame(
  gene = top500_genes,
  p_value = sapply(top500_genes, function(g) {
    fit <- aov(expr_staged[g, ] ~ factor(stage_info$stage))
    summary(fit)[[1]]["Pr(>F)"][1, 1]
  })
)
```

```{r}
#Multiple testing correction

anova_results$p_adj <- p.adjust(anova_results$p_value, method = "BH")
anova_results$significant <- anova_results$p_adj < 0.05

cat("Significant genes (FDR < 0.05):", sum(anova_results$significant), "\n")
print(anova_results %>% arrange(p_adj) %>% head(10))
cat("\n")
```

```{r}
#Heatmap Vizualization
#Out of the 500 most variable, only 85 are significant -> therefore, should only see small clustering....

top500_matrix <- z_scores_staged[top500_genes, stage_info$sample_id]

annotation <- data.frame(
  Stage = factor(stage_info$stage),
  row.names = stage_info$sample_id
)

pheatmap(
  top500_matrix,
  annotation_col = annotation,
  show_rownames = FALSE,
  show_colnames = FALSE,
  main = "Top 500 Variable Genes by Stage"
)
```

```{r}
#Identify the oncogenes

oncogenes <- c("MYC","KRAS","NRAS","HRAS","EGFR","ERBB2","PIK3CA",
               "BRAF","AKT1","AKT2","MDM2","CCND1","CDK4","CDK6",
               "MET","RET","ALK","ROS1","FGFR1","FGFR2")

oncogenes_in_data <- intersect(rownames(expr_staged), oncogenes)

cat("Oncogenes found:", length(oncogenes_in_data), "\n")
cat(paste(oncogenes_in_data, collapse = ", "), "\n\n")
```

```{r}
#Oncogene stage correlation
#Spearman correlation is a monotonic relationship between 2 variables
#In this case, we are looking at does expression go up or down with stage

if (length(oncogenes_in_data) > 0) {
 
  # Convert stage to numeric (I=1, II=2, III=3, IV=4)
  stage_numeric <- as.numeric(factor(
    stage_info$stage,
    levels = c("STAGE I", "STAGE II", "STAGE III", "STAGE IV")
  ))
 
  # Test correlation for each oncogene
  oncogene_results <- data.frame(
    gene = oncogenes_in_data,
    correlation = sapply(oncogenes_in_data, function(g) {
      cor.test(stage_numeric, expr_staged[g, ], method = "spearman")$estimate
    }),
    p_value = sapply(oncogenes_in_data, function(g) {
      cor.test(stage_numeric, expr_staged[g, ], method = "spearman")$p.value
    })
  )
 
  oncogene_results$p_adj <- p.adjust(oncogene_results$p_value, method = "BH")
  oncogene_results$significant <- oncogene_results$p_adj < 0.05
  oncogene_results$direction <- ifelse(oncogene_results$correlation > 0, "UP", "DOWN")
 
  cat("Oncogene correlation results:\n")
  print(oncogene_results %>% arrange(p_adj))
  cat("\n")
}
```
```{r}
#Plot significant outcomes

if (exists("oncogene_results")) {
  sig_oncogenes <- oncogene_results %>%
    filter(significant) %>%
    pull(gene)
 
  if (length(sig_oncogenes) > 0) {
    for (gene in sig_oncogenes) {
      corr <- oncogene_results$correlation[oncogene_results$gene == gene]
      p_adj <- oncogene_results$p_adj[oncogene_results$gene == gene]
     
      plot_data <- data.frame(
        expression = expr_staged[gene, ],
        stage = factor(stage_info$stage)
      )
     
      p <- ggplot(plot_data, aes(stage, expression, fill = stage)) +
        geom_boxplot() +
        labs(
          title = paste(gene, "| ρ =", round(corr, 3),
                       "| p_adj =", format(p_adj, scientific = TRUE, digits = 2)),
          x = "Tumor Stage",
          y = "Log2 Expression"
        ) +
        theme_bw() +
        theme(legend.position = "none")
     
      print(p)
    }
  } else {
    cat("No significant oncogenes to plot\n")
  }
}
```


